OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-06-25 paper/paper.Rmd
Cite USEARCH

<!--
- **"AND, BUT, THEREFORE" structure. start paragraph with question, end with why we should care. transitions to move the story along.**

- Go back through https://peerj.com/articles/1487/ to get all the arguments for why the qiime based methods are problematic…
1. The threshold with closed reference is a radius and with de novo its a diameter
2. The gg reference was defined for full length sequences, within a subregion there are duplicate and nearly duplicate sequences
3. Because VSEARCH/USEARCH is a greedy heuristic, its looking for a reference that is close. Although the ordering of the database is not relevant (it was with USEARCH) it gets the wrong number of its (see Figure 5)
-->

@@

**TODO**

@@

There are two main categories of OTU clustering algorithms: _de novo_ and
reference-based.
OptiClust is a _de novo_ clustering method which uses the distance score between
all pairs of sequences in the dataset to cluster them into OTUs by maximizing
the Matthews Correlation Coefficient (MCC) [@westcott_opticlust_2017].
This approach takes into account the distances between all pairs of sequences
when assigning query sequences to OTUs, in contrast to other _de novo_
algorithms such as VSEARCH which only considers the distance between the query
sequence and a representative centroid sequence in the OTU [**TODO cite vsearch and schloss papers, also maybe qiime**].
A limitation of _de novo_ clustering is that different OTU assignments will be
produced when new sequences are added to a dataset, making it difficult to use
_de novo_ clustering to compare OTUs between different studies.
Furthermore, since _de novo_ clustering requires calculating and comparing
distances between all sequences in a dataset, the execution time can be slow for
very large datasets.
Reference clustering attempts to overcome the limitations of _de novo_
clustering methods by using a representative set of sequences from a database,
with each reference sequence seeding an OTU.
Commonly, the Greengenes set of representative full length sequences clustered
at 97% similarity is used as the reference with VSEARCH
[@desantis_greengenes_2006; **TODO cite vsearch**].
Query sequences are then assigned to OTUs based on their similarity to the
reference sequences.
Any query sequences that are not within the distance threshold to any of the
reference sequences are either thrown out (closed reference clustering) or
clustered _de novo_ to create additional OTUs (open reference clustering).
While reference-based clustering is generally fast, it is limited by the
diversity of the reference database.
Rare or novel sequences in the sample will be lost in closed reference mode if
they are not represented by a similar sequence in the database.
**TODO: closed uses radius of 3%. de novo uses diameter of 3%. open uses both. vsearch is order dependent.**
Previous studies found that the OptiClust _de novo_ clustering algorithm created
the highest quality OTU assignments of all clustering methods [@westcott_opticlust_2017].

@@

We implemented the analysis workflow in Snakemake [@koster_snakemake_2012] and
wrote scripts in R [@r_core_team_r_2020], Python [@van_rossum_python_2009], and
GNU bash [@noauthor_bash_nodate].
Software used includes mothur v1.45.0 [@schloss_introducing_2009], VSEARCH
v2.13.3 [@rognes_vsearch_2016], numpy [@harris_array_2020], the tidyverse
metapackage [@wickham_welcome_2019], R Markdown [@xie_r_2018], ggtext
[@wilke_ggtext_2020], the SRA toolkit [@noauthor_sra-tools_nodate], and the
conda environment manager [@noauthor_anaconda_2016].
The complete workflow, manuscript, and conda environment are available at
**TODO: UPDATED REPO LINK**.

>>>>

<!--
- Go back through https://peerj.com/articles/1487/ to get all the arguments for why the qiime based methods are problematic…
1. The threshold with closed reference is a radius and with de novo its a diameter
2. The gg reference was defined for full length sequences, within a subregion there are duplicate and nearly duplicate sequences
3. Because VSEARCH/USEARCH is a greedy heuristic, its looking for a reference that is close. Although the ordering of the database is not relevant (it was with USEARCH) it gets the wrong number of its (see Figure 5)
-->

@@

**TODO.**
**"AND, BUT, THEREFORE" structure. start paragraph with question, end with why we should care. transitions to move the story along.**

@@

There are two main categories of OTU clustering algorithms: _de novo_ and
reference-based.
OptiClust is a _de novo_ clustering method which uses the distance score between
all pairs of sequences in the dataset to cluster them into OTUs by maximizing
the Matthews Correlation Coefficient (MCC) [@westcott_opticlust_2017].
This approach takes into account the distances between all pairs of sequences
when assigning query sequences to OTUs, in contrast to other _de novo_
methods such as the greedy clustering algorithms implemented by USEARCH and
VSEARCH, which only considers the distance between the query sequence and a
representative centroid sequence in the OTU [@edgar_search_2010;
@rognes_vsearch_2016].
A limitation of _de novo_ clustering is that different OTU assignments will be
produced when new sequences are added to a dataset, making it difficult to use
_de novo_ clustering to compare OTUs between different studies.
Furthermore, since _de novo_ clustering requires calculating and comparing
distances between all sequences in a dataset, the execution time can be slow for
very large datasets.
Reference clustering attempts to overcome the limitations of _de novo_
clustering methods by using a representative set of sequences from a database,
with each reference sequence seeding an OTU.
Commonly, the Greengenes set of representative full length sequences clustered
at 97% similarity is used as the reference with VSEARCH
[@desantis_greengenes_2006; @rognes_vsearch_2016; @noauthor_clustering_nodate].
Query sequences are then assigned to OTUs based on their similarity to the
reference sequences.
Any query sequences that are not within the distance threshold to any of the
reference sequences are either thrown out (closed reference clustering) or
clustered _de novo_ to create additional OTUs (open reference clustering).
While reference-based clustering is generally fast, it is limited by the
diversity of the reference database.
Rare or novel sequences in the sample will be lost in closed reference mode if
they are not represented by a similar sequence in the database.
**TODO: closed uses radius of 3%. de novo uses diameter of 3%. open uses both. vsearch is order dependent.**
Previous studies found that the OptiClust _de novo_ clustering algorithm created
the highest quality OTU assignments of all clustering methods [@westcott_opticlust_2017].

@@

We implemented the analysis workflow in Snakemake [@koster_snakemake_2012] and
wrote scripts in R [@r_core_team_r_2020], Python [@van_rossum_python_2009], and
GNU bash [@noauthor_bash_nodate].
Software used includes mothur v1.45.0 [@schloss_introducing_2009], VSEARCH
v2.13.3 [@rognes_vsearch_2016], numpy [@harris_array_2020], the tidyverse
metapackage [@wickham_welcome_2019], R Markdown [@xie_r_2018], ggtext
[@wilke_ggtext_2020], the SRA toolkit [@noauthor_sra-tools_nodate], and the
conda environment manager [@noauthor_anaconda_2016].
The complete workflow, manuscript, and conda environment are available at
https://github.com/SchlossLab/Sovacool_OptiFit_2021.
<!-- TODO: update repo link with journal title -->
